\documentclass[]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{url}
\usepackage{listings}
\usepackage[bookmarks]{hyperref}
\usepackage[a4paper, total={7in, 8in}]{geometry}

\lstset{basicstyle=\footnotesize\ttfamily,breaklines=true}
\lstloadlanguages{bash,Python}

\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\Tr}{Tr}
\newcommand{\realevol}{\texttt{realevol}}
\newcommand{\bra}[1]{\ensuremath{\langle#1|}}
\newcommand{\ket}[1]{\ensuremath{|#1\rangle}}
\newcommand{\aver}[1]{\ensuremath{\langle#1\rangle}}

\title{Lanczos short time propagation solver: implementation notes}
\author{Igor Krivenko}

\begin{document}
\maketitle

The Lanczos short time propagation solver (\realevol) is designed to solve models of many-body quantum systems with interacting fermions and bosons out of thermodynamical equilibrium. It is based on TRIQS library version 3.0.x. Similarly to other TRIQS applications, \realevol\ is written in C++17 and comes with a Python interface, which allows to run physical simulations by writing simple Python scripts.

Besides TRIQS, \realevol\ has the following prerequisites:
\begin{itemize}
    \item Boost C++ libraries, version 1.53 or newer (headers only);
    \item A working MPI implementation;
    \item ARPACK-NG version 3.6.0 or newer (\url{https://github.com/opencollab/arpack-ng});
\end{itemize}
The step-by-step installation instructions are given in the \texttt{README.md} file in the root source directory.

In \realevol\ Python scripts, time dependence of operators can be specified in 
two ways.
\begin{itemize}
    \item Using \texttt{TExpr} objects that are constructed from strings
    containing mathematical expressions (functions of variable $t$).
    A backend for this feature is provided by the C++ Mathematical Expression
    Parsing And Evaluation Library, a.k.a. ExprTk from Arash Partow
    (\url{https://github.com/ArashPartow/exprtk}). A copy of ExprTk is bundled
    to the source codes.

    \item Using \texttt{TInterp} objects, which are linear interpolators 
    constructed on a time mesh from a user-provided \texttt{NumPy} array of 
    complex values.
\end{itemize}

\section{\label{sec:formulation}Problem formulation}
\subsection{\label{sec:formulation:space}State space}

Let us consider a quantum system of many identical particles. Its state space $\mathcal{H}$ (Fock space) is generated by fixed finite sets of fermionic creation operators $\{c^\dagger_1,\ldots,c^\dagger_{N_f}\}$ and bosonic creation operators $\{b^\dagger_1,\ldots,b^\dagger_{N_b}\}$. The total space $\mathcal{H}$ factorizes as
\begin{equation}
    \mathcal{H} = \mathcal{H}_f \otimes \mathcal{H}_b,
\end{equation}
where $\mathcal{H}_f$ is finite-dimensional fermionic states space, $\dim(\mathcal{H}_f) = 2^{N_f}$ (as follows from Pauli exclusion principle). The bosonic state space $\mathcal{H}_b$ is formally infinitely dimensional, as the number of bosons created by every operator $b^\dagger_i$ is unbounded.
\realevol\ is an exact-diagonalization type solver, which is bound to work with vector spaces of finite dimensions. Therefore, one has to artificially truncate $\mathcal{H}_b$ so that its dimension is finite. Truncation is done by setting a cutoff on the number of bosonic excitations in each mode,
\begin{equation}
    b^\dagger_i\ket{n_i^\mathrm{max}} = 0, \quad i=1,\ldots,N_b.
\end{equation}
The truncation constants $n_i^\mathrm{max}$ are chosen by user as one of input parameters. For technical reasons, the user is asked to provide a list of positive numbers called \texttt{bits\_per\_boson} and defined such that
\begin{equation}
    n_i^\mathrm{max} = 2^\mathtt{bits\_per\_boson[{\it i}]} - 1.
\end{equation}

According to a standard TRIQS notation, the abstract index $i$ must take a form
of a pair \texttt{(block\_index,inner\_index)} in Python scripts. Each element
of the pair can either be an integer, or a string. The rationale behind this
notation is that Green's function containers share the same type of indices in TRIQS. In many physical problems Green's functions turn out to be block-diagonal. This is why there must be one index to address a diagonal block,
and another index to choose a column/row within that block. For example, local
dynamics of a 3-orbital quantum impurity without spin-orbital coupling can be
described by a GF with block indices \texttt{'up'}, \texttt{'down'} (spin
projections) and inner indices \texttt{1}, \texttt{2}, \texttt{3} (orbitals). Therefore, there are 6
index combinations used to label operators $c^\dagger$/$c$:
\texttt{('up',1)}, \texttt{('up',2)}, \texttt{('up',3)}, \texttt{('down',1)},
\texttt{('down',2)} and \texttt{('down',3)}.

\subsection{\label{sec:formulation:initial}Initial state}
At the initial time moment ($t=0$) the system is prepared in a known state characterized by density matrix $\hat\rho_0$. \realevol\ supports three
qualitatively different types of initial states.

\begin{enumerate}
    \item Pure state, $\hat\rho_0 = \ket{\psi}\bra{\psi}$.

    Instead of specifying $\ket{\psi}$ as a list of coefficients w.r.t. some basis in $\mathcal{H}$, user is expected to provide a generating operator $\hat P$,
    \begin{equation}
        \ket{\psi} = \frac{\hat P\ket{0}}
                    {\sqrt{\bra{0}\hat P^\dagger \hat P\ket{0}}}.
    \end{equation}
    (A runtime error will occur if $\hat P\ket{0} = 0$).
    This approach is more convenient in practice, because it does not require knowledge about the particular choice of Fock basis in $\mathcal{H}$.

    \item Ground state(s) of a given Hamiltonian $\hat H_0$.
    \begin{equation}
        \hat\rho_0 = \frac{1}{g_\mathrm{gs}}\sum_{\psi\in\{\mathrm{gs}\}} \ket{\psi}\bra{\psi}.
    \end{equation}
    $g_\mathrm{gs}$ stands for the degeneracy degree of the ground state.

    \item Thermal state with a given Hamiltonian $\hat H_0$ at temperature $T$.
    \begin{equation}
        \hat\rho_0 = \frac{\exp(-\hat H_0/T)}{Z},
        \quad Z = \Tr[\exp(-\hat H_0/T)].
    \end{equation}
\end{enumerate}

Initial state (2) can be seen as the zero temperature limit of (3).

\subsection{\label{sec:formulation:observables}Observables}

Evolution of the system at $t>0$ is governed by a time-dependent Hamiltonian $\hat H(t)$.

There are currently three classes of time-dependent observables \realevol\ can compute \cite{NoneqDMFT}.
\begin{itemize}
    \item Expectation values
    \begin{equation}\label{1point_corr}
        \aver{A(t)} = \Tr[\hat\rho_0 \hat A(t)].
    \end{equation}
    
    \item Two-point correlation functions
    \begin{equation}\label{2point_corr}
        \aver{A(t) B(t')} = \Tr[\hat\rho_0 \hat A(t) \hat B(t')].
    \end{equation}
    
    \item Three-point correlation functions
    \begin{equation}\label{3point_corr}
    \aver{A(t) B(t') C(t'')} = \Tr[\hat\rho_0 \hat A(t) \hat B(t') \hat C(t'')].
    \end{equation}
\end{itemize}

It also provides a few convenience functions for computation of the most commonly used 2-point correlators.

\begin{itemize}
    \item Greater Green's function of fermions,
    \begin{equation}\label{greater_gf}
        G^>_{ij}(t,t') = -i\frac{1}{\hbar}\aver{c_i(t) c^\dagger_j(t')} =
        -i\frac{1}{\hbar}\Tr\left[\hat\rho_0 c_i(t) c^\dagger_j(t')\right].
    \end{equation}

    \item Lesser Green's function of fermions,
    \begin{equation}\label{lesser_gf}
        G^<_{ij}(t,t') = i\frac{1}{\hbar}\aver{c^\dagger_j(t') c_i(t)} =
        i\frac{1}{\hbar}\Tr\left[\hat\rho_0 c^\dagger_j(t') c_i(t)\right].
    \end{equation}

    \item Generalized susceptibility ($n_i = c^\dagger_i c_i$),
    \begin{equation}\label{susceptibility}
        \chi_{ij}(t,t') = -i\frac{1}{\hbar}\aver{n_i(t) n_j(t')} =
        -i\frac{1}{\hbar}\Tr\left[\hat\rho_0 n_i(t) n_j(t')\right].
    \end{equation}
\end{itemize}

Here, $c_i(t)$, $c^\dagger_i(t)$ and $n_i(t)$ are Heisenberg operators related to their Schr\"odinger counterparts as
\begin{equation}
    \hat O(t) = \hat U(0,t)\hat O \hat U(t,0).
\end{equation}
Operators $\hat A$, $\hat B$, $\hat C$ must be multivariate polynomials
of $c_i/c^\dagger_i$ and $b_i/b^\dagger_i$. Under this assumption, the multi-point correlation functions \ref{1point_corr}--\ref{3point_corr} can be split into a sum of correlators of monomials $c^\dagger\ldots c^\dagger c \ldots c b^\dagger \ldots b^\dagger b \ldots b$. \realevol\ performs the splitting procedure and summation over all contributions automatically, so in the rest of the text I will assume that $\hat A$, $\hat B$, $\hat C$ are monomials.

Following a standard derivation, one can show that the time evolution operator $\hat U(t,t')$ is given by a time-(anti)ordered exponential,
\begin{equation}
    \hat U(t,t') = \left\{
        \begin{array}{ll}
        \mathbb{T}_t\exp\left[
        -\frac{i}{\hbar}\int_{t'}^t d\tilde t\hat H(\tilde t) \right], &t>t',\\
        \mathbb{\bar T}_t\exp\left[
        -\frac{i}{\hbar}\int_{t'}^t d\tilde t\hat H(\tilde t) \right], &t<t'.
        \end{array}
    \right.
\end{equation}

\section{\label{sec:init_state}\texttt{InitState}: Representation of the initial state}

Any initial state $\hat\rho_0$ supported by \realevol\ can be represented as a finite sum of weighted pure states.
\begin{equation}
    \hat\rho_0 = \sum_m w_m \ket{\psi_m}\bra{\psi_m},\quad\sum_m w_m = 1.
\end{equation}
The way state vectors $\ket{\psi_m}$ and their respective weights $w_m$ are chosen, varies between the three cases listed in \ref{sec:formulation:initial} and will be discussed below.
At this point I only require that $\ket{\psi_m}$ are normalized and orthogonal for different $m$, $\langle\psi_m|\psi_k\rangle = \delta_{mk}$.
Before I proceed with this discussion, I need to introduce the notion of state space partition.

The full state space of the problem can, in general, be partitioned into a direct sum of subspaces,
\begin{equation}
    \mathcal{H} = \mathcal{H}_1 \oplus \mathcal{H}_2 \oplus \ldots \oplus
                  \mathcal{H}_S,\quad 1\le S\le \dim(\mathcal{H}).
\end{equation}
If a given state $\ket{\psi}\in\mathcal{H}_s$, it is possible to save memory by storing only expansion coefficients of $\ket{\psi}$ over the basis vectors of
$\mathcal{H}_s$. This optimization technique is implemented in \texttt{InitState}, the Python object used to store $\hat\rho_0$ in \realevol\
(the corresponding C++ type is called \texttt{init\_state}). An instance of
\texttt{InitState} contains information about space partition ($S$, and a list of basis vectors for each $\mathcal{H}_s$) and a list of all contributing state/weight pairs accompanied by indices of the respective subspaces.
\begin{equation}
    \mathtt{InitState} = \{\mathcal{H}_s\}, \{(\ket{\psi_m}, w_m, s)\}.
\end{equation}

\texttt{InitState}, as well as functions to create initial states, are found in \texttt{realevol.init\_state} Python module.

\section{\label{sec:make_pure_init_state}Function \texttt{make\_pure\_init\_state()}}

Python function \texttt{make\_pure\_init\_state()} creates a pure initial state (see Sec.~\ref{sec:formulation:initial}, point 1). It takes four arguments.
\begin{itemize}
    \item A Python object of type \texttt{realevol.operators\_texpr.Operator} or
    \texttt{realevol.operators\_tinterp.Operator}
    representing the generating operator $\hat P$.

    $\hat P$ must be time-independent and such that $\hat P\ket{0} \neq 0$.
    \item Index set for fermions, i.e. a \texttt{set} of \texttt{(block\_index,inner\_index)} pairs to label all operators $c^\dagger_i$/$c_i$.

    \item Index set for bosons, i.e. a \texttt{set} of \texttt{(block\_index,inner\_index)} pairs to label all operators $b^\dagger_i$/$b_i$.

    \item \texttt{bits\_per\_boson}, number of bits per bosonic
    degree of freedom passed as Python dictionary\\
    \texttt{\{(block\_index,inner\_index):int\}}. Must be empty if
    system contains no bosons.

\end{itemize}

The two index sets and \texttt{bits\_per\_boson} together define the full Fock
space $\mathcal{H}$ (see Sec.~\ref{sec:formulation:space}).

\texttt{make\_pure\_init\_state()} returns an instance of \texttt{InitState}
containing one triplet
\begin{equation}
    \left(
        \frac{\hat P\ket{0}}{\sqrt{\bra{0}\hat P^\dagger \hat P\ket{0}}},
        1,
        1
    \right),
\end{equation}
and two subspaces, $\mathcal{H}=\mathcal{H}_1\oplus\mathcal{H}_2$.
$\mathcal{H}_1$ is spanned by all basis vectors of $\mathcal{H}$ that have non-zero overlaps with $\hat P\ket{0}$. The rest of the basis vectors span
$\mathcal{H}_2$.

\section{\label{sec:make_equilibrium_init_state}Function \texttt{make\_equilibrium\_init\_state()}}

As its name suggests, function \texttt{make\_equilibrium\_init\_state()} creates
an equilibrium initial state with a given (possibly zero) temperature $T$ (Sec.~\ref{sec:formulation:initial}, points 2 and 3). The function has the following arguments.
\begin{itemize}
    \item $\hat H_0$, Hamiltonian of the system at $t=0$ as a
    \texttt{realevol.operators\_texpr.Operator} or\\
    \texttt{realevol.operators\_tinterp.Operator} object.
    \item Index set for fermions.
    \item Index set for bosons.
    \item Temperature $T$.
    \item Dictionary of calculation parameters (will be explained shortly).
    \item \texttt{bits\_per\_boson}, number of bits per bosonic
    degree of freedom.
\end{itemize}

In order to construct $\hat\rho_0$, one needs to (partially) diagonalize $\hat H_0$ and obtain some of its eigenpairs $(E_n,\ket{\psi_n})$ with the smallest energies. The diagonalization is preceded by a preparatory step, where $\mathcal{H}$
is partitioned into a direct sum of invariant subspaces of $\hat H_0$,
\begin{equation}
    \mathcal{H} = \mathcal{H}_1 \oplus \mathcal{H}_2 \oplus \ldots \oplus
        \mathcal{H}_S,\quad \hat H_0\mathcal{H}_s = \mathcal{H}_s.
\end{equation}
This sort of space partition effectively transforms $\hat H_0$ into a block-diagonal form and greatly reduces computational costs of the diagonalization. \realevol~employs the first phase of so called autopartition
algorithm described in Sec. 4.2 of \cite{CTHYB}.

Diagonalization of each diagonal block is then performed with one of three
methods, depending on the dimension of the corresponding $\mathcal{H}_s$.
\begin{itemize}
    \item $\dim(\mathcal{H}_s) = 1$, diagonalization of a $1\times1$ matrix is trivial.
    \item $\dim(\mathcal{H}_s)$ is smaller than the value of calculation parameter \texttt{arpack\_min\_matrix\_size}.

    LAPACK routine \texttt{zheev} is called to obtain all eigenpairs of the diagonal block.

    \item $\dim(\mathcal{H}_s)$ is greater than or equal to \texttt{arpack\_min\_matrix\_size}.

    ARPACK is used to compute a few states with the lowest energies within
    the block. Calculation parameter \texttt{arpack\_tolerance} can be used
    to control the eigenvalue convergence tolerance. Another relevant parameter
    is \texttt{arpack\_ncv}. It is a dictionary \texttt{\{int:int\}} that allows to set NCV (number of Lanczos vectors) for individual subspaces (keys of the dictionary are subspace indices $s$).
\end{itemize}

\texttt{make\_equilibrium\_init\_state()} fills \texttt{InitState} with
eigenvectors of $\hat H_0$ and their statistical weights,
\begin{equation}
    \mathtt{InitState} = \{\mathcal{H}_s\},
    \left\{\left(\ket{\psi_n}, \frac{\exp(-E_n/T)}{Z}, s\right)\right\}.
\end{equation}

The total number of included eigenstates depends on the cutoff imposed on
the statistical weight $\exp(-E_n/T)/Z$. This cutoff can be adjusted via
\texttt{min\_rel\_weight} parameter and equals the machine epsilon by default.
It is worth noting that usually only a few subspaces contribute to the
initial state if the temperature is not too high. If $T=0$, then the cutoff
parameter is ignored, and only the orthogonalized ground states of $\hat H_0$ are included,
\begin{equation}
    \mathtt{InitState} = \{\mathcal{H}_s\},
    \left\{\left(\ket{\psi^\mathrm{gs}_n}, \frac{1}{g_\mathrm{gs}}, s\right)\right\}.
\end{equation}


The amount of information being printed out by \texttt{make\_equilibrium\_init\_state()} is controlled via the \texttt{verbosity} parameter.

\section{\label{sec:worldlines}World line decomposition of two-time correlation functions}

Definitions of the correlation functions listed in Sec. \ref{sec:formulation:observables} involve products of Schr\"odinger operators and evolution operators $\hat U(t, t')$,
\begin{equation}\label{prods}
\begin{aligned}
    \hat A(t) &= \hat U(0,t)\hat A \hat U(t,0),\\
    \hat A(t) \hat B(t') &= \hat U(0,t)\hat A\hat U(t,t')\hat B \hat U(t',0),\\
    \hat A(t) \hat B(t') \hat C(t'') &= \hat U(0,t)\hat A\hat U(t,t')\hat B \hat U(t',t'') \hat C \hat U(t'',0).
\end{aligned}
\end{equation}

Proceeding along the lines of Sec.~\ref{sec:init_state}, I construct a partition of space $\mathcal{H}$,
\begin{equation}\label{tilde_partition}
    \mathcal{H} =
    \mathcal{\tilde H}_1 \oplus \mathcal{\tilde H}_2
    \oplus \ldots \oplus
    \mathcal{\tilde H}_{\tilde S},\quad 1\le \tilde S\le \dim(\mathcal{H}),
\end{equation}
so that
\begin{enumerate}
    \item $\mathcal{\tilde H}_s$ are invariant subspaces of $\hat H(t)$ for all $t>0$, $\hat H(t)\mathcal{\tilde H}_s = \mathcal{\tilde H}_s$;
    \item Fermionic creation and annihilation operators establish one-to-one mappings between subspaces, $c_i\mathcal{\tilde H}_s = \mathcal{\tilde H}_p$, $c^\dagger_i\mathcal{\tilde H}_s = \mathcal{\tilde H}_p$ (these equalities also cover the case when $c_i/c^\dagger_i\ket{\psi} = 0$ for some $\ket{\psi}\in\mathcal{\tilde H}_s$).
\end{enumerate}
Once again, the optimal partition fulfilling these two criteria is constructed
by running both phases of the autopartition algorithm (Sec. 4.2 of \cite{CTHYB}).

It follows from the criterion 1 that the evolution operator is block-diagonal, $\hat U(t,t')\mathcal{\tilde H}_s = \mathcal{\tilde H}_s$. Together with the criterion 2, this identity leads to a conclusion that products \ref{prods} can be calculated block-wise. Let superindices on operators denote their
blocks w.r.t. space partition \ref{tilde_partition}.
\begin{equation}
\begin{aligned}
    (A(t))^{s_1s_2} &=
    \hat U(0,t)^{s_1s_1}\hat A^{s_1s_2}
    \hat U(t,0)^{s_2s_2},\\
    (\hat A(t) \hat B(t'))^{s_1s_3} &=
    \hat U(0,t)^{s_1s_1}\hat A^{s_1s_2}
    \hat U(t,t')^{s_2s_2}\hat B^{s_2s_3}
    \hat U(t',0)^{s_3s_3},\\
    (\hat A(t) \hat B(t') \hat C(t''))^{s_1s_4} &=
    \hat U(0,t)^{s_1s_1} \hat A^{s_1s_2}
    \hat U(t,t')^{s_2s_2} \hat B^{s_2s_3}
    \hat U(t',t'')^{s_3s_3} \hat C^{s_3s_4}
    \hat U(t'',0)^{s_4s_4}.
\end{aligned}
\end{equation}

To conclude calculation of the $P$-point correlation functions, one needs to do averaging over the initial state,
\begin{equation}
    \Tr[\hat\rho_0 \hat O_P] =
    \sum_{m}w_m\Tr\left[\ket{\psi_m}\bra{\psi_m}\hat O_P\right] =
    \sum_{m}w_m\bra{\psi_m}\hat O_P\ket{\psi_m}.
\end{equation}
with $\hat O_P$ being one of the products from \ref{prods}.

My next step is to project states $\ket{\psi_m}$ onto the subspaces $\mathcal{\tilde H}_s$.
In general, space partition generated by $\hat H(t)$ is different from the one used to represent $\hat\rho_0$, that is $\mathcal{H}_s$. This difference typically occurs when symmetry of the system changes at $t=0$ or at a later time. It, therefore, is necessary to compute ``branching'' of subspaces: Find out what subspaces $\mathcal{\tilde H}_s$ contain vectors forming the basis set of a given subspace $\mathcal{H}_s$.
Knowing branching information, it is easy to project $\ket{\psi_m}$ onto $\mathcal{\tilde H}_s$ by picking relevant amplitudes from $\ket{\psi_m}$ and
forming a new vector out of them.

\begin{equation}\label{1point_wl}
    \aver{A(t)} =
    \sum_m w_m\sum_{s_1s_2}
    \bra{\psi_m}
    \hat U(0,t)^{s_1s_1}\hat A^{s_1s_2}
    \hat U(t,0)^{s_2s_2}
    \ket{\psi_m}.
\end{equation}
\begin{equation}\label{2point_wl}
    \aver{A(t) B(t')} =
    \sum_m w_m\sum_{s_1s_3}
    \bra{\psi_m}
    \hat U(0,t)^{s_1s_1}\hat A^{s_1s_2}
    \hat U(t,t')^{s_2s_2}\hat B^{s_2s_3}
    \hat U(t',0)^{s_3s_3}
    \ket{\psi_m}.
\end{equation}
\begin{equation}\label{3point_wl}
    \aver{A(t) B(t') C(t'')} =
    \sum_m w_m\sum_{s_1s_4}
    \bra{\psi_m}
    \hat U(0,t)^{s_1s_1} \hat A^{s_1s_2}
    \hat U(t,t')^{s_2s_2} \hat B^{s_2s_3}
    \hat U(t',t'')^{s_3s_3} \hat C^{s_3s_4}
    \hat U(t'',0)^{s_4s_4}
    \ket{\psi_m}.
\end{equation}

I call the combination $(m; s_1,\ldots,s_{P+1})$ a world line contributing to a $P$-point correlation function. Since matrix multiplication scales as $O(N^3)$ with multipliers' size, splitting the trace into world lines and summing over them makes for a major performance improvement. Indeed, many zero matrix elements, whose summation would otherwise increase computation time, are left out by this procedure.

\section{\label{sec:evol} State evolution in real time}

As it follows from equations \ref{1point_wl}, \ref{2point_wl} and \ref{3point_wl}, states projected onto a subspace $\mathcal{\tilde H}_s$ and propagated in time are basic building blocks making up world lines,
\begin{equation}\label{Uss_psi}
    \hat U(t,t')^{ss}\ket{\psi} =
    \left\{
    \begin{array}{ll}
    \mathbb{T}_t\exp\left[
    -\frac{i}{\hbar}\int_{t'}^t d\tilde t\hat H(\tilde t)^{ss} \right]\ket{\psi}, &t>t',\\
    \mathbb{\bar T}_t\exp\left[
    -\frac{i}{\hbar}\int_{t'}^t d\tilde t\hat H(\tilde t)^{ss} \right]\ket{\psi}, &t<t'.
    \end{array}
    \right.
\end{equation}
(Note the block indices on the Hamiltonian).

In \realevol, all observables and time-dependent operators are defined on a uniform time grid $t_n = n\Delta t, t_n\in[0;t_\mathrm{max}]$. There are several supported ways to discretize equation \ref{Uss_psi}. First of all, if all matrix elements of $\hat H(t)^{ss}$ happen to be time-independent, discretization is trivial,
\begin{equation}\label{U_psi_static}
    \hat U(t_n,t_m')^{ss}\ket{\psi} =
    \exp\left[-\frac{i}{\hbar}(t_n-t_m')\hat H^{ss} \right]\ket{\psi}.
\end{equation}
Otherwise, $\hat U(t_n,t_m')^{ss}$ is constructed as a product of evolution operators over short time steps $\Delta t$ (\textit{short time propagation}),
\begin{equation}
    \hat U(t_n,t_m')^{ss}\ket{\psi} =
    \left\{
    \begin{array}{ll}
        \hat U(t_n,t_n-\Delta t)^{ss}\ldots
        \hat U(t_m'+2\Delta t,t_m'+\Delta t)^{ss}
        \hat U(t_m'+\Delta t,t_m')^{ss}\ket{\psi}, &t>t',\\
        \hat U(t_n,t_n+\Delta t)^{ss}\ldots
        \hat U(t_m'-2\Delta t,t_m'-\Delta t)^{ss}
        \hat U(t_m'-\Delta t,t_m')^{ss}\ket{\psi}, &t<t'.
    \end{array}
\right.
\end{equation}

The time step $\Delta t$ is assumed to be short compared to all characteristic time scales of the system in question. If this condition is fulfilled, one can approximate integrals in \ref{Uss_psi} using quadrature rules, i.e. construct interpolating Hamiltonians.
\begin{equation}\label{U_psi_dynamic}
    \hat U(t_n + \Delta t, t_n)\ket{\psi} \approx
    \exp\left[-\frac{i}{\hbar}\Delta t \hat H_\mathrm{int}^{ss}(t_n) \right]
    \ket{\psi}.
\end{equation}

\realevol\ implements three approximations for the time integral (switchable at run time).

\begin{enumerate}
    \item Rectangle rule,
    \begin{equation}\label{U_rectangle}
        \hat H_\mathrm{int}^{ss}(t_n) = \hat H^{ss}(t_n + \Delta t/2).
    \end{equation}
    \item Trapezoid rule,
    \begin{equation}\label{U_trapezoid}
        \hat H_\mathrm{int}^{ss}(t_n) =
        \frac{\hat H^{ss}(t_n) + \hat H^{ss}(t_n + \Delta t)}{2}.
    \end{equation}
    \item Simpson rule,
    \begin{equation}\label{U_simpson}
        \hat H_\mathrm{int}^{ss}(t_n) =
        \frac{\hat H^{ss}(t_n) + 4\hat H^{ss}(t_n + \Delta t/2) +
        \hat H^{ss}(t_n + \Delta t)}{6}.
    \end{equation}
\end{enumerate}
As it is seen from the definitions above, the interpolation rules differ in computation cost, as they require 1, 2, and 3 evaluations of $\hat H^{ss}$ respectively.

Now comes the main and the most expensive part of the algorithm, calculation of the propagated state \ref{U_psi_dynamic}/\ref{U_psi_static}. Given $\mathcal{\tilde H}_s$, \realevol\ chooses one of the following propagation techniques:
\begin{itemize}
    \item If $\dim(\mathcal{\tilde H}_s) = 1$, exponentials in \ref{U_psi_dynamic}/\ref{U_psi_static} are scalars, which makes propagation trivial.

    \item $\dim(\mathcal{H}_s)$ is moderate (below a user-defined limit).

    The matrix exponential is constructed via LAPACK diagonalization of $\hat H_\mathrm{int}^{ss}$. In case $\hat H^{ss}$ is static (eq. \ref{U_psi_static}), the diagonalization is done only once for all time arguments of $\hat U(t_n, t'_m)^{ss}$.

    \item $\dim(\mathcal{H}_s)$ is above the user-defined limit.

    The r.h.s. of \ref{U_psi_dynamic} is computed using the Lanczos algorithm
    as described in \cite{matexp,Potthoff}. {\it Since the Lanczos algorithm reveals only a part of the spectrum, the short time propagation has to be used even if $\hat H^{ss}$ is static}.
\end{itemize}

\section{\label{sec:operators_texpr}Python representation of time-dependent 
operators}

Representation of quantum-mechanical operators in Python scripts strives to be as close to the mathematical notation as possible. It is implemented in two modules,
\begin{lstlisting}[language=Python]
from realevol.texpr import TExpr          # Time-dependent expressions
from realevol.operators_texpr import *    # Second quantization operators
\end{lstlisting}
\texttt{TExpr} is a small utility class made to encapsulate mathematical expressions involving the time variable $t$. Upon creation, \texttt{TExpr} takes a string and tries to interpret it as an expression. After \texttt{TExpr} object is created, one can call it with a real number as an argument, and it will return value of the expression with $t$ substituted by the argument.
\begin{lstlisting}[language=Python]
>>> te = TExpr("2*cos(2*pi*t)")
>>> te(0.5) # Shall return 2*cos(2*pi*0.5)
(-2+0j)
\end{lstlisting}

Operators are constructed as algebraic expressions made of creation, annihilation and occupation number operator objects.
\begin{lstlisting}[language=Python]
>>> cd1 = c_dag('up',0)     # cd1 creates fermion ('up',0)
>>> cd1
1*C^+(up,0)
>>> c1 = c('up',0)          # c1 destroys fermion ('up',0)
>>> c1
1*C(up,0)
>>> n1 = n('up',0)          # n1 is the number of fermions ('up',0)
>>> n1
1*C^+(up,0)C(up,0)
>>> bd1 = a_dag('B',0)      # bd1 creates boson ('B',0)
>>> bd1
1*A^+(B,0)
>>> b1 = a('B',0)           # b1 destroys boson ('B',0)
>>> b1
1*A(B,0)
\end{lstlisting}
\textit{N.B.: There is no occupation number operator for bosons.}

The elementary operators are combined into algebraic expressions with coefficients being real/complex constants or \texttt{TExpr} objects.
\begin{lstlisting}[language=Python]
>>> H1 = -1.0 * (n('up',0) + n('dn',0)) + 2.0 * n('up',0) * n('dn',0)
>>> H1
-1*C^+(dn,0)C(dn,0) + -1*C^+(up,0)C(up,0) + 2*C^+(dn,0)C^+(up,0)C(up,0)C(dn,0)
>>> H2 = TExpr("0.1*cos(0.5*t)") * (a_dag('B',0) + a('B',0))
>>> H2
0.1*cos(0.5*t)*A^+(B,0) + 0.1*cos(0.5*t)*A(B,0)
\end{lstlisting}

\section{\label{sec:operators_tinterp}Python representation of time-dependent 
operators: alternative approach}

An alternative to the \texttt{TExpr} object is the linear interpolator object
\texttt{TInterp}.

\begin{lstlisting}[language=Python]
from triqs.gf import MeshReTime             # Real time mesh
from realevol.tinterp import TInterp        # Linear interpolator
from realevol.operators_tinterp import *    # Second quantization operators
from numpy import array, cos, pi
\end{lstlisting}

\texttt{TInterp} instances are constructed from a given time mesh and a 
\texttt{NumPy} array of values. If all values in the array happen to be equal,
the resulting object will internally store only one number and return it upon
every evaluation call.
\begin{lstlisting}[language=Python]
>>> mesh = MeshReTime(0, 5.0, 101)
>>> ti = TInterp(mesh, array([2*cos(2*pi*t) for t in mesh]))
>>> ti(0.99) # Shall return linearly interpolated value of 2*cos(2*pi*0.99)
(1.9804226065180615+0j)
>>> 2*cos(2*pi*0.99) # Exact value
(1.9960534568565431+0j)
\end{lstlisting}
The constant-valued instances can also be constructed explicitly from the 
corresponding real/complex constant.
\begin{lstlisting}[language=Python]
>>> ti_const = TInterp(7+5j)
>>> ti_const(1)
(7+5j)
\end{lstlisting}

The use of \texttt{TInterp} coefficients in operator expressions is analogous
to that of \texttt{TExpr}.

\textit{N.B.: \texttt{TExpr}-based operators and \texttt{TInterp}-based 
operators cannot be mixed within one expression.}

Evaluation speed of \texttt{TInterp} objects is normally much higher 
compared to \texttt{TExpr}, but there is a trade-off between evaluation 
accuracy and the amount of memory required to store data arrays with interpolation coefficients.

\section{\label{sec:compute}\texttt{compute\_*()} functions}

Computation of the 1-, 2- and 3-point correlators is performed by the following Python functions defined in the module \texttt{realevol.realevol}.

\begin{enumerate}

    \item Time-dependent expectation value $\aver{A(t)}$ (\ref{1point_corr}).
    \begin{lstlisting}[language=Python]
    compute_expectval(A, init_state, h, t_mesh, params)
    \end{lstlisting}
    Arguments:
    \begin{itemize}
        \item \texttt{A}: Operator $\hat{A}$ (must be time-independent).
        \item \texttt{init\_state}: Initial state of the system at $t=0$ as an \texttt{InitState} object (see Sec.~\ref{sec:init_state} for more detail.)
        \item \texttt{h}: Time-dependent Hamiltonian of the system $\hat H(t)$.
        \item \texttt{t\_mesh}: Real time mesh to compute the expectation value on (\texttt{triqs.gf.MeshReTime}).
        \item \texttt{params}: Various computation parameters, see below.
    \end{itemize}
    Returns a scalar-valued \texttt{triqs.gf.gf.Gf} object defined on the 1-dimensional real time mesh (\texttt{t\_mesh}).

    \item 2-point correlator $\aver{A(t) B(t)}$ (\ref{2point_corr}).
    \begin{lstlisting}[language=Python]
        compute_correlator_2t(A, B, init_state, h, t_mesh, params)
    \end{lstlisting}
    Arguments:
    \begin{itemize}
        \item \texttt{A}: Operator $\hat{A}$ (must be time-independent).
        \item \texttt{B}: Operator $\hat{B}$ (must be time-independent).
        \item \texttt{init\_state}: Initial state of the system at $t=0$ as an \texttt{InitState} object (see Sec.~\ref{sec:init_state} for more detail.)
        \item \texttt{h}: Time-dependent Hamiltonian of the system $\hat H(t)$.
        \item \texttt{t\_mesh}: Real time mesh to compute the correlator on (\texttt{triqs.gf.MeshReTime}).
        \item \texttt{params}: Various computation parameters, see below.
    \end{itemize}
    Returns a scalar-valued \texttt{triqs.gf.gf.Gf} object defined on 2-dimensional Cartesian product of the real time mesh with itself (\texttt{t\_mesh} $\times$ \texttt{t\_mesh}).

    \item 3-point correlator $\aver{A(t) B(t) C(t)}$ (\ref{3point_corr}).
    \begin{lstlisting}[language=Python]
        compute_correlator_3t(A, B, C, init_state, H, t_mesh, params)
    \end{lstlisting}
    Arguments:
    \begin{itemize}
        \item \texttt{A}: Operator $\hat{A}$ (must be time-independent).
        \item \texttt{B}: Operator $\hat{B}$ (must be time-independent).
        \item \texttt{C}: Operator $\hat{C}$ (must be time-independent).
        \item \texttt{init\_state}: Initial state of the system at $t=0$ as an \texttt{InitState} object (see Sec.~\ref{sec:init_state} for more detail.)
        \item \texttt{h}: Time-dependent Hamiltonian of the system $\hat H(t)$.
        \item \texttt{t\_mesh}: Real time mesh to compute the correlator on (\texttt{triqs.gf.MeshReTime}).
        \item \texttt{params}: Various computation parameters, see below.
    \end{itemize}
    Returns a scalar-valued \texttt{triqs.gf.gf.Gf} object defined on 3-dimensional Cartesian product of the real time mesh with itself (\texttt{t\_mesh} $\times$ \texttt{t\_mesh} $\times$ \texttt{t\_mesh}).
    
    \item Greater Green's function of fermions $G^>_{ij}(t,t')$ (\ref{greater_gf}).
    \begin{lstlisting}[language=Python]
        compute_g_g(gf_struct, init_state, h, t_mesh, params)
    \end{lstlisting}
    Arguments:
    \begin{itemize}
        \item \texttt{gf\_struct}: Green's function structure.
        \item \texttt{init\_state}: Initial state of the system at $t=0$ as an \texttt{InitState} object (see Sec.~\ref{sec:init_state} for more detail.)
        \item \texttt{h}: Time-dependent Hamiltonian of the system $\hat H(t)$.
        \item \texttt{t\_mesh}: Real time mesh to compute the GF on (\texttt{triqs.gf.MeshReTime}).
        \item \texttt{params}: Various computation parameters, see below.
    \end{itemize}
    Returns a \texttt{triqs.gf.block\_gf.BlockGf} object with matrix-valued blocks defined on 2-dimensional Cartesian product of the real time mesh with itself (\texttt{t\_mesh} $\times$ \texttt{t\_mesh}).

    \item Lesser Green's function of fermions $G^<_{ij}(t,t')$ (\ref{lesser_gf}).
    \begin{lstlisting}[language=Python]
        compute_g_l(gf_struct, init_state, h, t_mesh, params)
    \end{lstlisting}
    Arguments:
    \begin{itemize}
        \item \texttt{gf\_struct}: Green's function structure.
        \item \texttt{init\_state}: Initial state of the system at $t=0$ as an \texttt{InitState} object (see Sec.~\ref{sec:init_state} for more detail.)
        \item \texttt{h}: Time-dependent Hamiltonian of the system $\hat H(t)$.
        \item \texttt{t\_mesh}: Real time mesh to compute the GF on (\texttt{triqs.gf.MeshReTime}).
        \item \texttt{params}: Various computation parameters, see below.
    \end{itemize}
    Returns a \texttt{triqs.gf.block\_gf.BlockGf} object with matrix-valued blocks defined on 2-dimensional Cartesian product of the real time mesh with itself (\texttt{t\_mesh} $\times$ \texttt{t\_mesh}).

    \item Generalized susceptibility $\chi_{ij}(t,t')$ (\ref{susceptibility}).
    \begin{lstlisting}[language=Python]
        compute_chi(chi_indices, init_state, h, t_mesh, params)
    \end{lstlisting}
    Arguments:
    \begin{itemize}
        \item \texttt{chi\_indices}: Indices of components of the generalized susceptibility to compute.
        \item \texttt{init\_state}: Initial state of the system at $t=0$ as an \texttt{InitState} object (see Sec.~\ref{sec:init_state} for more detail.)
        \item \texttt{h}: Time-dependent Hamiltonian of the system $\hat H(t)$.
        \item \texttt{t\_mesh}: Real time mesh to compute the susceptibility on (\texttt{triqs.gf.MeshReTime}).
        \item \texttt{params}: Various computation parameters, see below.
    \end{itemize}
    Returns a matrix-valued \texttt{triqs.gf.gf.Gf} object defined on 2-dimensional Cartesian product of the real time mesh with itself (\texttt{t\_mesh} $\times$ \texttt{t\_mesh}). The dimension of the matrix is equal to the size of \texttt{chi\_indices}.

\end{enumerate}

Operator-valued arguments of these functions can be either \texttt{realevol.operators\_texpr.Operator}\linebreak or
 \texttt{realevol.operators\_tinterp.Operator} objects, but not both mixed in the same function call. \texttt{gf\_struct} defines the structure of the greater/lesser Green's functions to be measured. It has a form of a list, \texttt{[(block\_index, [inner\_indices])]} (see Sec.~ \ref{sec:formulation:space}). \texttt{chi\_indices} controls what components of the susceptibility are to be measured. It is a \texttt{set} of
 \texttt{(block\_index,inner\_index)} pairs. 
{\it Important}: All indices met in \texttt{gf\_struct}, \texttt{chi\_indices} and in \texttt{h} must be known to the \texttt{InitState} object used in the simulation.

\texttt{params} is a parameters dictionary with the following elements.
\begin{itemize}
    \item \texttt{verbosity}, positive integer, verbosity level;
    \item \texttt{hbar}, floating point, Planck's constant, 1.0 by default;
    \item \texttt{hamiltonian\_interpol}, $\hat H(t)^{ss}$ interpolation type (Sec.~\ref{sec:evol}), one of \texttt{'Rectangle'}, \texttt{'Trapezoid'}, \texttt{'Simpson'};
    \item \texttt{t\_ranges}. For the $P$-point correlation function, a list of $P$ pairs \texttt{(t\_min,t\_max)}. Only combinations of time arguments lying within the specified ranges will be calculated. By default all the ranges are $(-\infty,\infty)$. (Green's functions and susceptibilities are special cases of $P = 2$).
    \item \texttt{delta\_t\_max}. For the $P$-point correlation function, a list of $P-1$ floats. The $p$-th elements of this list determines the maximal time separation of the $p$-th and $(p+1)$-th time argument for the corresponding contribution to be calculated. By default all the floats are set to $\infty$. (Green's functions and susceptibilities are special cases of $P = 2$).
    \item \texttt{lanczos\_min\_matrix\_size}, positive integer, minimum $\dim(\mathcal{\tilde H}_s)$ that triggers use of the Lanczos algorithm for state propagation (Sec.~\ref{sec:evol});
    \item \texttt{lanczos\_gs\_energy\_tol}, dictionary \texttt{\{int:float\}}, allows to set Lanczos eigenvalue convergence tolerance for individual subspaces $\mathcal{\tilde H}_s$. Generation of Krylov basis vectors stops when the lowest eigenvalue of the tridiagonal Lanczos matrix converges within that tolerance;
    \item \texttt{lanczos\_max\_krylov\_dim}, dictionary \texttt{\{int:float\}}, allows to set a cap on the Krylov basis size for individual subspaces $\mathcal{\tilde H}_s$.
\end{itemize}

\section{\label{sec:example}Simulation example}

The following Python script is to be run as \texttt{python3 example.py} or
\texttt{mpirun -n \$NPROCS python3 example.py}.

\hrule

\lstinputlisting[language=python]{example.py}

\begin{thebibliography}{9}

\bibitem{NoneqDMFT}
``Nonequilibrium dynamical mean-field theory and its applications''.
H. Aoki, N. Tsuji, M. Eckstein, M. Kollar, T. Oka, and Ph. Werner,
Rev. Mod. Phys. {\bf 86}, 779 (2014).

\bibitem{CTHYB}
``TRIQS/CTHYB: A continuous-time quantum Monte Carlo hybridisation
expansion solver for quantum impurity problems''.
P. Seth, I. Krivenko, M. Ferrero and O. Parcollet,
Comp. Phys. Comm. {\bf 200}, 274-284 (2016).

\bibitem{matexp}
``The Krylov Subspace Methods for the Computation of Matrix Exponentials''.
H. Wang,
PhD thesis, University of Kentucky, 2015.

\bibitem{Potthoff}
``Krylov-space approach to the equilibrium and nonequilibrium
single-particle Green's function''.
M. Balzer, N. Gdaniec and M. Potthoff,
J. Phys.: Condens. Matter {\bf 24}, 035603 (2012).

\end{thebibliography}

\end{document}
